---
title: "Biodiversity Indicators"
output: # html_notebook
editor_options: 
  chunk_output_type: console
---

# R Set-up

```{r}
# Clear workspace
rm(list = ls())

# Libraries 
library(terra)
library(here)
library(purrr)
library(dplyr)
library(tigris)
library(sf)

# Authenticate aws
source(here::here('aws_cred.R'))

# Define directories
dir <- here()
dat.dir <- file.path(dir, "data", "raw")
roi.dir <- file.path(dir, "data", "studyarea")
dum.dir <- file.path(dir, "data", "dummy")

```

We developed biodiversity indicators in three categories:  Special Places, Species & Habitats, and Grassland Condition. 


# 0.1 Study Area

Load in the study areas. Will align all data to this crs.

ROI 1 = TBD
ROI 2 = Clip any data to this one
ROI 3 = TBD

```{r}

roi1 <- vect(file.path(roi.dir, "class_study_area_aligned_cnvc_Dec2025.shp"))
roi2 <- vect(file.path(roi.dir, "modelling_extent_aligned_cnvc_Jan2026.shp"))
roi3 <- vect(file.path(roi.dir, "modelling_study_area_aligned_cnvc_Jan2026.shp"))

# plot(roi1)
# plot(roi2)
# plot(roi3, main="roi3")

```


# 0.2 Dummy Data

```{r}
base_dum <- rast(file.path(dum.dir, "eccc_reclass_reproj_algn_binary.tif"))
fut_dum <- rast(file.path(dum.dir, "AAFC_LUTS_2020_5-classes.tif"))

plot(base_dum)
plot(fut_dum)
```


# 1.0 Special Places

## 1.1 Key biodiversity areas

Includes Accepted and draft KBAS => combine into one
Check for data sensitivities


```{r}
# Define path to the geodatabase(s)
kba_path <- file.path(dat.dir, "1_special_places", "1.1_KBAs", "AcceptedKBAs.gdb")
kba_draft_path <- file.path(dat.dir, "1_special_places", "1.1_KBAs", "DRAFTKBAs.gdb")

# List all layers in the geodatabase(s)
st_layers(kba_path)
st_layers(kba_draft_path)

# Read the main feature class from the geodatabase(s)
kba <- st_read(kba_path, layer = "KBASite")
kba_draft <- st_read(kba_draft_path, layer = "DRAFT_KBASite")

# Convert sf objects to terra SpatVector objects
kba <- vect(kba)
kba_draft <- vect(kba_draft)

# Harmonize to Study Area CRS
kba <- project(kba_draft, roi1)
kba_draft <- project(kba_draft, roi1)

# clean-up
rm(kba_path, kba_draft_path)
```

## 1.2 ECCC priority places for species-at-risk (SAR)

```{r}
# Define path to the geodatabase(s)
priority_path <- file.path(dat.dir, "1_special_places", "1.2_PriorityPlaces", "PriorityPlaces.gdb")

# List all layers in the geodatabase(s)
st_layers(priority_path)

# Read the main feature class from the geodatabase(s)
priority <- st_read(priority_path, layer = "PriorityPlacesBoundary")

# Convert sf objects to terra SpatVector objects
priority <- vect(priority)

# Harmonize to Study Area CRS
priority <- project(priority, roi1)

# clean-up
rm(priority_path)
```

## 1.3 ECCC community nominated priority places

```{r}
# Define path to the geodatabase(s)
comm_priority_path <- file.path(dat.dir, "1_special_places", "1.3_CommunityNominated_PriorityPlaces", "CommunityNominated_PriorityPlaces.gdb")

# List all layers in the geodatabase(s)
st_layers(comm_priority_path)

# Read the main feature class from the geodatabase(s)
comm_priority <- st_read(comm_priority_path, layer = "CommunityNominated_PriorityPlaces")

# Convert sf objects to terra SpatVector objects
comm_priority <- vect(comm_priority)

# Harmonize to Study Area CRS
comm_priority <- project(comm_priority, roi1)

# clean-up
rm(comm_priority_path)
```

## 1.4 Relatively rare grasses

Waiting for data from Alison


# 2.0 Species & Habitats

## 2.1 Climate Refugia

Multiple scenarios
2050s
2080s

Each timeframe has a layer for songbirds, trees, and combined for a total of 6.

```{r, eval=F}
# Define path to the geodatabase(s)
refugia_path <- file.path(dat.dir, "2_species_and_habitats", "2.1_clim_refugia", "NA_Refugia", "2050s")

# List all layers in the geodatabase(s)
st_layers(refugia_path)

# Read the main feature class from the geodatabase(s)
comm_priority <- st_read(refugia_path, layer = "CommunityNominated_PriorityPlaces")

# Convert sf objects to terra SpatVector objects
comm_priority <- vect(comm_priority)

# Harmonize to Study Area CRS
comm_priority <- project(comm_priority, roi1)

# clean-up
rm(refugia_path)
```

## 2.2 Upland priority areas for Prairie Habitat 

Add details

```{r}
# Define path to the file(s)
upland_path <- file.path(dat.dir, "2_species_and_habitats", "2.2_PriorityAreaAnalysisResults", "PriorityAreasPolygons", "UplandSpecies", "X1_Landbirds.ABF_M_q80.shp")

upland <- vect(upland_path)

# Harmonize to Study Area CRS
upland <- project(upland, roi1)

# clean-up
rm(upland_path)

```


## 2.3 Areas important for bird population growth

```{r}
# Define path to the file(s)
bird_pop_path <- file.path(dat.dir, "2_species_and_habitats", "2.3_CGBWG_Data_Share", "PriorityAreasPolygons", "UplandSpecies", "X1_Landbirds.ABF_M_q80.shp")
```


## 2.4 Richness of species-at-risk

Struggling to download (Ronnie to send his copy)

## 2.5.1 Wildlife habitat availability on farmland: High-Very High Capacity

```{r}
# Define path to the geodatabase(s)
farm_path <- file.path(dat.dir, "species_and_habitats", "farmland_habitat", "AEI_MGMNT_WHC_GDB", "AEI_MGMNT_WHC_GDB.gdb")

# List all layers in the geodatabase(s)
st_layers(farm_path)

# Read the main feature class from the geodatabase(s)
comm_priority <- st_read(refugia_path, layer = "CommunityNominated_PriorityPlaces")

# Convert sf objects to terra SpatVector objects
comm_priority <- vect(comm_priority)

# Harmonize to Study Area CRS
comm_priority <- project(comm_priority, roi1)

# clean-up
rm(refugia_path)
```

## 2.5.2 Wildlife habitat availability on farmland: Large-Moderate Decrease

```{r}

```

## 2.6 Species distribution model for grassland birds

Analysis being done externally by Barry Robinson

## 2.7 Habitat suitability for grassland species

Analysis being done externally by James Paterson


# 3.0 Grassland Condition

## 3.1 Intact grasslands

```{r}

```

## 3.2 Connectivity

```{r}

```

## 3.3 Native grasslands

```{r}

```

