---
title: "Biodiversity Indicators"
output: # html_notebook
editor_options: 
  chunk_output_type: console
---

# R Set-up

```{r}
# Clear workspace
rm(list = ls())

# Libraries 
library(terra)
library(here)
library(purrr)
library(dplyr)
library(tigris)
library(sf)

# Authenticate aws
source(here::here('aws_cred.R'))

# Base S3 location (GDAL-style path)
s3_base <- "/vsis3/sagemaker-gst-stage.sharing/ciara-hovis/TomoPrairie/bd_data"

dat.dir <- file.path(s3_base, "raw")
roi.dir <- file.path(s3_base, "studyarea")
dum.dir <- file.path(s3_base, "dummy")
ind.dir <- file.path(s3_base, "indicators")

# Define directories (OLD)
# dir <- here()
# dat.dir <- file.path(dir, "data", "raw")
# roi.dir <- file.path(dir, "data", "studyarea")
# dum.dir <- file.path(dir, "data", "dummy")
```

We developed biodiversity indicators in three categories:  Special Places, Species & Habitats, and Grassland Condition. 


# 0.1 Study Area

Load in the study areas. Will align all data to this crs.

ROI 1 = TBD
ROI 2 = Clip any data to this one
ROI 3 = TBD

```{r}

roi1 <- vect(file.path(roi.dir, "class_study_area_aligned_cnvc_Dec2025.shp"))
roi2 <- vect(file.path(roi.dir, "modelling_extent_aligned_cnvc_Jan2026.shp"))
roi3 <- vect(file.path(roi.dir, "modelling_study_area_aligned_cnvc_Jan2026.shp"))

# plot(roi1, main="roi1")
# plot(roi2, main="roi2")
# plot(roi3, main="roi3")

```



# 1.0 Special Places

## 1.1 Key biodiversity areas

Includes Accepted and draft KBAS => combine into one
Check for data sensitivities

```{r}
# Define path to the geodatabase(s)
kba_path <- file.path(dat.dir, "1_special_places", "1.1_KBAs", "AcceptedKBAs.gdb")
kba_draft_path <- file.path(dat.dir, "1_special_places", "1.1_KBAs", "DRAFTKBAs.gdb")

# List all layers in the geodatabase(s)
st_layers(kba_path)
st_layers(kba_draft_path)

# Read the main feature class from the geodatabase(s)
kba <- st_read(kba_path, layer = "KBASite")
kba_draft <- st_read(kba_draft_path, layer = "DRAFT_KBASite")

# Convert sf objects to terra SpatVector objects
kba <- vect(kba)
kba_draft <- vect(kba_draft)

# Harmonize to Study Area CRS
kba <- project(kba, roi3)
kba_draft <- project(kba_draft, roi3)

# clean-up
rm(kba_path, kba_draft_path)
gc()

```

## 1.2 ECCC priority places for species-at-risk (SAR)

```{r}
# Define path to the geodatabase(s)
priority_path <- file.path(dat.dir, "1_special_places", "1.2_PriorityPlaces", "PriorityPlaces.gdb")

# List all layers in the geodatabase(s)
st_layers(priority_path)

# Read the main feature class from the geodatabase(s)
priority <- st_read(priority_path, layer = "PriorityPlacesBoundary")

# Convert sf objects to terra SpatVector objects
priority <- vect(priority)

# Harmonize to Study Area CRS
priority <- project(priority, roi3)

# clean-up
rm(priority_path)
gc()

```

## 1.3 ECCC community nominated priority places

```{r}
# Define path to the geodatabase(s)
comm_priority_path <- file.path(dat.dir, "1_special_places", "1.3_CommunityNominated_PriorityPlaces", "CommunityNominated_PriorityPlaces.gdb")

# List all layers in the geodatabase(s)
st_layers(comm_priority_path)

# Read the main feature class from the geodatabase(s)
comm_priority <- st_read(comm_priority_path, layer = "CommunityNominated_PriorityPlaces")

# Convert sf objects to terra SpatVector objects
comm_priority <- vect(comm_priority)

# Harmonize to Study Area CRS
comm_priority <- project(comm_priority, roi3)

# clean-up
rm(comm_priority_path)
gc()

```

## 1.4 Relatively rare grasses

Need protected area data & ecoregion(?) data
```{r}
# --- Load data ---
eco_reg <- vect(file.path(
  roi.dir,
  "national-ecological-framework-for-canada",
  "Ecoregions",
  "ecoregions_clean_prairies.shp"
))

prot_area <- st_read(
  file.path(
    dat.dir,
    "1_special_places",
    "1.4_ProtectedAreas",
    "ProtectedConservedArea_2025",
    "ProtectedConservedArea_2025.gdb"
  ),
  layer = "ProtectedConservedArea_2025"
)


# --- Keep polygon geometries only and convert to terra ---
prot_area_vect <- vect(
  prot_area[st_geometry_type(prot_area) %in% c("POLYGON", "MULTIPOLYGON"), ]
)

# --- Align CRS and extent ---
eco_reg <- project(eco_reg, roi3)
prot_area_vect <- crop(project(prot_area_vect, roi3), roi3)

# --- Dissolve all protected areas into a single layer ---
prot_union <- aggregate(prot_area_vect, dissolve = TRUE)

# --- Compute UNPROTECTED area (ecoregions minus protected areas) ---
unprot_all <- erase(eco_reg, prot_union)

# --- Optional: split into list by ecoregion (same structure as your original output) ---
eco_ids <- unique(unprot_all$ECOREGION_)

unprot_list <- lapply(eco_ids, function(id) {
  aggregate(
    unprot_all[unprot_all$ECOREGION_ == id, ],
    by = "ECOREGION_",
    dissolve = TRUE
  )
})

names(unprot_list) <- eco_ids

rare1 <- unprot_list[[1]]
rare2 <- unprot_list[[2]]
rare3 <- unprot_list[[3]]
rare4 <- unprot_list[[4]]
rare5 <- unprot_list[[5]]
rare6 <- unprot_list[[6]]
rare7 <- unprot_list[[7]]

rm(eco_reg, prot_area, prot_area_vect, prot_union, unprot_all, eco_ids, unprot_list)
gc()

```


# 2.0 Species & Habitats

## 2.1 Climate Refugia

Multiple scenarios
2050s

Not doing trees b/c trees aren't really a priority for grasslands


```{r, eval=T}
# Define path to the geodatabase(s)
refugia_bird_path <- file.path(dat.dir, "2_species_and_habitats", "2.1_clim_refugia", "NA_Refugia", "2050s", "RCP85_Songbirds", "HabitatGCMCombo", "Grassland_songbird_2050refugia_combo85.tif")

refugia_bird <- rast(refugia_bird_path)

# Harmonize to Study Area CRS
crs(refugia_bird) <- "+proj=laea +lat_0=45 +lon_0=-100 +datum=WGS84 +units=m +no_defs"

refugia_bird <- project(refugia_bird, roi3)

# Function to convert raster values to percentiles (0–1)
to_percentile <- function(r) {
  vals <- values(r, na.rm = TRUE)
  ecdf_fun <- ecdf(vals)
  
  r_pct <- r
  values(r_pct) <- ecdf_fun(values(r))
  
  return(r_pct)
}

# calc percentiles
bird_pct <- to_percentile(refugia_bird)

bird_pct <- crop(bird_pct, roi3, mask=T)

# threshold at 80th percentile
bird_top <- bird_pct >= 0.9

refugia_bird_poly <- as.polygons(bird_top, dissolve = TRUE, na.rm = TRUE)

# clean-up
rm(refugia_bird_path, refugia_bird, bird_pct, bird_top)
gc()

```

## 2.2 Upland priority areas for Prairie Habitat 


```{r}
# Define path to the file(s)
upland_path <- file.path(dat.dir, "2_species_and_habitats", "2.2_PriorityAreaAnalysisResults", "PriorityAreasPolygons", "UplandSpecies", "X1_Landbirds.ABF_M_q80.shp")

upland <- vect(upland_path)

# Harmonize to Study Area CRS
upland <- project(upland, roi3)

# clean-up
rm(upland_path)

```


## 2.3 Areas important for bird population growth

Paused

```{r, eval=F}
# Define path to the file(s)
bird_pop_path <- file.path(dat.dir, "2_species_and_habitats", "2.3_CGBWG_Data_Share", "PriorityAreasPolygons", "UplandSpecies", "X1_Landbirds.ABF_M_q80.shp")

# clean-up
rm(bird_pop_path)

```


## 2.4 Richness of species-at-risk

Need to re-work to use gdb

```{r, eval=F}
# Define path 
sar_path <- file.path(dat.dir, "2_species_and_habitats", "2.4_EBAR_SAR_Heatmap", "HeatmapEcoshapeOverview.shp")

sar <- vect(sar_path)

# Harmonize to Study Area CRS
sar <- project(sar, roi3)

# clean-up
rm(sar_path)

```


## 2.5.1 Wildlife habitat availability on farmland: High-Very High Capacity

```{r, eval=F}
# Define path to the geodatabase(s)
farm_path <- file.path(dat.dir, "species_and_habitats", "farmland_habitat", "AEI_MGMNT_WHC_GDB", "AEI_MGMNT_WHC_GDB.gdb")

# List all layers in the geodatabase(s)
st_layers(farm_path)

# Read the main feature class from the geodatabase(s)
comm_priority <- st_read(refugia_path, layer = "CommunityNominated_PriorityPlaces")

# Convert sf objects to terra SpatVector objects
comm_priority <- vect(comm_priority)

# Harmonize to Study Area CRS
comm_priority <- project(comm_priority, roi3)

# clean-up
rm(refugia_path)
```

## 2.5.2 Wildlife habitat availability on farmland: Large-Moderate Decrease

```{r}

```

## 2.6 Species distribution model for grassland birds

Analysis being done externally by Barry Robinson

## 2.7 Habitat suitability for grassland species

Analysis being done externally by James Paterson


# 3.0 Grassland Condition

## 3.1 Intact grasslands

```{r, eval=F}
# Define path to the file(s)
intact_path <- file.path(dat.dir, "3_grass_condition", "3.1_WWF_Plowprint_2024", "Spatial_data", "WWF_PP_2024_report_full_extent", "WWF_PP_2024_report_full_extent.tif")

intact <- rast(intact_path)

# Harmonize to Study Area CRS
intact <- project(intact, roi3)

# clean-up
rm(intact_path)

```

## 3.2 Connectivity

```{r}

```

## 3.3 Native grasslands

```{r}

```


# 4.0 Combine indicator data

```{r}
# special places
ind_sp <- c(kba, kba_draft, 
            priority, 
            comm_priority,
            rare1, rare2, rare3, rare4, rare5, rare6, rare7)

# species and habitats
ind_sh <- c(refugia_bird_poly,
            upland)

# grassland condition
ind_gc <- c()


```

Crop to roi
```{r}


crop_svc_to_roi <- function(svc_obj, roi) {
  svc(
    lapply(as.list(svc_obj), function(x) {
      out <- intersect(crop(makeValid(x), roi), roi)
      out <- do.call(rbind, as.list(out))  # works whether out is list/svc/vector
      if (is.null(out) || nrow(out) == 0) return(NULL)
      out
    }) |> Filter(Negate(is.null), x = _)
  )
}



ind_sp_crop <- crop_svc_to_roi(ind_sp, roi3)
ind_sh_crop <- crop_svc_to_roi(ind_sh, roi3)

```


Save to s3
```{r}
numbers_sp <- c("1_1a", "1_1b", "1_2", "1_3",
                "1_4a", "1_4b", "1_4c", "1_4d", "1_4e", "1_4f", "1_4g")

stopifnot(length(numbers_sp) == length(ind_sp_crop))

for (i in seq_along(ind_sp_crop)) {

  fname <- paste0("ind_sp_", numbers_sp[i], ".gpkg")

  writeVector(
    ind_sp_crop[[i]],
    file.path(ind.dir, fname),
    overwrite = TRUE
  )
}

numbers_sh <- c("2_1","2_2")

stopifnot(length(numbers_sh) == length(ind_sh_crop))

for (i in seq_along(ind_sh_crop)) {

  fname <- paste0("ind_sp_", numbers_sh[i], ".gpkg")

  writeVector(
    ind_sh_crop[[i]],
    file.path(ind.dir, fname),
    overwrite = TRUE
  )
}

```

