---
title: "Biodiversity Indicators"
output: # html_notebook
editor_options: 
  chunk_output_type: console
---

# R Set-up

```{r}
# Clear workspace
rm(list = ls())

# Libraries 
library(terra)
library(here)
library(purrr)
library(dplyr)
library(tigris)
library(sf)

# Authenticate aws
source(here::here('aws_cred.R'))

# Base S3 location (GDAL-style path)
s3_base <- "/vsis3/sagemaker-gst-stage.sharing/ciara-hovis/TomoPrairie/bd_data"

dat.dir <- file.path(s3_base, "raw")
roi.dir <- file.path(s3_base, "studyarea")
dum.dir <- file.path(s3_base, "dummy")
ind.dir <- file.path(s3_base, "indicators")
tmp_dir <- "/home/sagemaker-user/WG_Biodiversity/tmp_data"

# Define directories (OLD)
# dir <- here()
# dat.dir <- file.path(dir, "data", "raw")
# roi.dir <- file.path(dir, "data", "studyarea")
# dum.dir <- file.path(dir, "data", "dummy")
```


We developed biodiversity indicators in three categories:  Special Places, Species & Habitats, and Grassland Condition. 


# 0.1 Study Area

Load in the study areas. Will align all data to this crs.

ROI 1 = TBD
ROI 2 = Clip any data to this one
ROI 3 = TBD

```{r}

roi1 <- vect(file.path(roi.dir, "class_study_area_aligned_cnvc_Dec2025.shp"))
roi2 <- vect(file.path(roi.dir, "modelling_extent_aligned_cnvc_Jan2026.shp"))
roi3 <- vect(file.path(roi.dir, "modelling_study_area_aligned_cnvc_Jan2026.shp"))

# plot(roi1, main="roi1")
# plot(roi2, main="roi2")
# plot(roi3, main="roi3")

```



# 1.0 Special Places

## 1.1 Key biodiversity areas

Includes Accepted and draft KBAS => combine into one
Check for data sensitivities

```{r}
# Define path to the geodatabase(s)
kba_path <- file.path(dat.dir, "1_special_places", "1.1_KBAs", "AcceptedKBAs.gdb")
kba_draft_path <- file.path(dat.dir, "1_special_places", "1.1_KBAs", "DRAFTKBAs.gdb")

# List all layers in the geodatabase(s)
st_layers(kba_path)
st_layers(kba_draft_path)

# Read the main feature class from the geodatabase(s)
kba <- st_read(kba_path, layer = "KBASite")
kba_draft <- st_read(kba_draft_path, layer = "DRAFT_KBASite")

# Convert sf objects to terra SpatVector objects
kba <- vect(kba)
kba_draft <- vect(kba_draft)

# Harmonize to Study Area CRS
kba <- project(kba, roi3)
kba_draft <- project(kba_draft, roi3)

# clean-up
rm(kba_path, kba_draft_path)
gc()

```

## 1.2 ECCC priority places for species-at-risk (SAR)

```{r}
# Define path to the geodatabase(s)
priority_path <- file.path(dat.dir, "1_special_places", "1.2_PriorityPlaces", "PriorityPlaces.gdb")

# List all layers in the geodatabase(s)
st_layers(priority_path)

# Read the main feature class from the geodatabase(s)
priority <- st_read(priority_path, layer = "PriorityPlacesBoundary")

# Convert sf objects to terra SpatVector objects
priority <- vect(priority)

# Harmonize to Study Area CRS
priority <- project(priority, roi3)

# clean-up
rm(priority_path)
gc()

```

## 1.3 ECCC community nominated priority places

```{r}
# Define path to the geodatabase(s)
comm_priority_path <- file.path(dat.dir, "1_special_places", "1.3_CommunityNominated_PriorityPlaces", "CommunityNominated_PriorityPlaces.gdb")

# List all layers in the geodatabase(s)
st_layers(comm_priority_path)

# Read the main feature class from the geodatabase(s)
comm_priority <- st_read(comm_priority_path, layer = "CommunityNominated_PriorityPlaces")

# Convert sf objects to terra SpatVector objects
comm_priority <- vect(comm_priority)

# Harmonize to Study Area CRS
comm_priority <- project(comm_priority, roi3)

# clean-up
rm(comm_priority_path)
gc()

```

## 1.4 Relatively rare grasses

Need protected area data & ecoregion(?) data
```{r}
# --- Load data ---
eco_reg <- vect(file.path(
  roi.dir,
  "national-ecological-framework-for-canada",
  "Ecoregions",
  "ecoregions_clean_prairies.shp"
))

prot_area <- st_read(
  file.path(
    dat.dir,
    "1_special_places",
    "1.4_ProtectedAreas",
    "ProtectedConservedArea_2025",
    "ProtectedConservedArea_2025.gdb"
  ),
  layer = "ProtectedConservedArea_2025"
)



# --- Keep polygon geometries only and convert to terra ---
prot_area_vect <- vect(
  prot_area[st_geometry_type(prot_area) %in% c("POLYGON", "MULTIPOLYGON"), ]
)

# --- Align CRS and extent ---
eco_reg <- project(eco_reg, roi3)
prot_area_vect <- crop(project(prot_area_vect, roi3), roi3)

# --- Dissolve all protected areas into a single layer ---
prot_union <- aggregate(prot_area_vect, dissolve = TRUE)

# --- Compute UNPROTECTED area (ecoregions minus protected areas) ---
unprot_all <- erase(eco_reg, prot_union)

# --- Optional: split into list by ecoregion (same structure as your original output) ---
eco_ids <- unique(unprot_all$ECOREGION_)

unprot_list <- lapply(eco_ids, function(id) {
  aggregate(
    unprot_all[unprot_all$ECOREGION_ == id, ],
    by = "ECOREGION_",
    dissolve = TRUE
  )
})

names(unprot_list) <- eco_ids

rare1 <- unprot_list[[1]]
rare2 <- unprot_list[[2]]
rare3 <- unprot_list[[3]]
rare4 <- unprot_list[[4]]
rare5 <- unprot_list[[5]]
rare6 <- unprot_list[[6]]
rare7 <- unprot_list[[7]]

rm(eco_reg, prot_area, prot_area_vect, prot_union, unprot_all, eco_ids, unprot_list)
gc()

```


# 2.0 Species & Habitats

## 2.1 Climate Refugia

Multiple scenarios
2050s

Not doing trees b/c trees aren't really a priority for grasslands

RCP85 scenario

Threshold is 95th percentile

```{r, eval=T}
# Define path to the geodatabase(s)
refugia_grassbird_path <- file.path(
  dat.dir, "2_species_and_habitats", "2.1_clim_refugia", "NA_Refugia", "2050s", "RCP85_Songbirds",
  "HabitatGCMCombo", "Grassland_songbird_2050refugia_combo85.tif"
)

refugia_scrubbird_path <- file.path(
  dat.dir, "2_species_and_habitats", "2.1_clim_refugia", "NA_Refugia", "2050s", "RCP85_Songbirds",
  "HabitatGCMCombo", "Scrub_songbird_2050refugia_combo85.tif"
)

# Read in data
refugia_grassbird <- rast(refugia_grassbird_path)
refugia_scrubbird <- rast(refugia_scrubbird_path)

# Harmonize to Study Area CRS
crs(refugia_grassbird) <- "+proj=laea +lat_0=45 +lon_0=-100 +datum=WGS84 +units=m +no_defs"
crs(refugia_scrubbird) <- "+proj=laea +lat_0=45 +lon_0=-100 +datum=WGS84 +units=m +no_defs"

refugia_grassbird <- project(refugia_grassbird, roi3)
refugia_scrubbird <- project(refugia_scrubbird, roi3)

# Crop/mask before percentile
refugia_grassbird <- crop(refugia_grassbird, roi3, mask = TRUE)
refugia_scrubbird <- crop(refugia_scrubbird, roi3, mask = TRUE)

# Function to convert raster values to percentiles (0–1)
to_percentile <- function(r) {
  vals <- values(r, na.rm = TRUE)
  ecdf_fun <- ecdf(vals)

  r_pct <- r
  values(r_pct) <- ecdf_fun(values(r))

  return(r_pct)
}

# Calculate percentiles
grassbird_pct <- to_percentile(refugia_grassbird)
scrubbird_pct <- to_percentile(refugia_scrubbird)

# Threshold at top percentiles
grassbird_top <- grassbird_pct >= 0.95
scrubbird_top <- scrubbird_pct >= 0.95

# Convert TRUE/FALSE → 1 / NA
grassbird_bin <- ifel(grassbird_top, 1, NA)
scrubbird_bin <- ifel(scrubbird_top, 1, NA)

# Convert to polygons (TRUE areas only)
refugia_grassbird_poly <- as.polygons(grassbird_bin, dissolve = TRUE, na.rm = TRUE)
refugia_scrubbird_poly <- as.polygons(scrubbird_bin, dissolve = TRUE, na.rm = TRUE)

# clean-up
rm(refugia_grassbird_path, refugia_scrubbird_path, refugia_grassbird, refugia_scrubbird,
   to_percentile, grassbird_pct, scrubbird_pct, grassbird_top, scrubbird_top)
gc()

```

## 2.2 Upland priority areas for Prairie Habitat 

```{r}
# Define path to the file(s)
upland_path <- file.path(dat.dir, "2_species_and_habitats", "2.2_PriorityAreaAnalysisResults", "PriorityAreasPolygons", "UplandSpecies", "X1_Landbirds.ABF_M_q80.shp")

upland <- vect(upland_path)

# Harmonize to Study Area CRS
upland <- project(upland, roi3)

# clean-up
rm(upland_path)

```

## 2.3 Richness of species-at-risk

Calculated percentiles for SAR and only keeping polygons in the 95th percentile

```{r, eval=T}
# Define path 
sar_path <- file.path(dat.dir, "2_species_and_habitats", "2.3_EBAR_SAR_Heatmap", "EBAR_SAR_Heatmap", "EBAR_SAR_Heatmap.gdb")

sar <- st_read(sar_path, layer = "HeatmapEcoshapeOverview")
sar <- vect(sar)

# Harmonize to Study Area CRS
sar <- project(sar, roi3)

# Crop/mask before percentile
sar <- crop(sar, roi3)

# Calc percentiles
sar$HeatmapRangeCount_pct <- rank(sar$HeatmapRangeCount, ties.method = "average") / 
                             sum(!is.na(sar$HeatmapRangeCount))

# Keep polygons at or above the 90th percentile
sar_top <- sar[sar$HeatmapRangeCount_pct >= 0.95, ]

# plot(sar_top, "HeatmapRangeCount")
# plot(sar_top, "HeatmapRangeCount_pct")
# plot(sar, "HeatmapRangeCount")
# plot(sar, "HeatmapRangeCount_pct")

# clean-up
rm(sar_path, sar)

```


## 2.4 Wildlife habitat availability on farmland: High-Very High Capacity

Areas with high and very high capacity (cap)

```{r, eval=T}
# Define path to the geodatabase(s)
farm_path <- file.path(dat.dir, "2_species_and_habitats", "2.4_farmland_habitat", "farmland_habitat", "AEI_MGMNT_WHC_GDB", "AEI_MGMNT_WHC.gdb")

# Read in data
farm_F <- st_read(farm_path, layer = "AEI_MGMNT_WHC_F_TREND") # feeding
farm_R <- st_read(farm_path, layer = "AEI_MGMNT_WHC_R_TREND") # reproduction

farm_F <- vect(farm_F)
farm_R <- vect(farm_R)

# Harmonize to Study Area CRS
farm_F <- project(farm_F, roi3)
farm_R <- project(farm_R, roi3)

# Extract areas of high and very high capacity
# WHC_R/F_2020_CLASS
# Very High = 1
# High = 2

farm_F_cap <- farm_F[farm_F$WHC_F_2020_CLASS <= 2, ]
farm_R_cap <- farm_R[farm_R$WHC_R_2020_CLASS <= 2, ]


# clean-up
rm(farm_path, farm_F, farm_R)
```

## 2.5 Wildlife habitat availability on farmland: Large-Moderate Decrease

Large and moderate decrease (chng)

```{r}
# Define path to the geodatabase(s)
farm_path <- file.path(dat.dir, "2_species_and_habitats", "2.4_farmland_habitat", "farmland_habitat", "AEI_MGMNT_WHC_GDB", "AEI_MGMNT_WHC.gdb")

# Read in data
farm_F <- st_read(farm_path, layer = "AEI_MGMNT_WHC_F_TREND") # feeding
farm_R <- st_read(farm_path, layer = "AEI_MGMNT_WHC_R_TREND") # reproduction

farm_F <- vect(farm_F)
farm_R <- vect(farm_R)

# Harmonize to Study Area CRS
farm_F <- project(farm_F, roi3)
farm_R <- project(farm_R, roi3)

# Extract areas of high and very high capacity
# WHC_R/F_00_20_CHG_CLASS
# Large dec = 5
# Mod dec = 4

farm_F_chng <- farm_F[farm_F$WHC_F_2020_CLASS >= 4, ]
farm_R_chng <- farm_R[farm_R$WHC_R_2020_CLASS >= 4, ]

# clean-up
rm(farm_path, farm_F, farm_R)
```

## 2.6 Species distribution model for grassland birds

Analysis being done externally by Barry Robinson

## 2.7 Habitat suitability for grassland species

Analysis being done externally by James Paterson

## 2.X Areas important for bird population growth

Paused, might drop at some point


# 3.0 Grassland Condition

## 3.1 Intact grasslands

Values 0 and 1 == intact

```{r, eval=T}
# Define path to the file(s)
intact_path <- file.path(dat.dir, "3_grass_condition", "3.1_WWF_Plowprint_2024", "Spatial_data", "WWF_PP_2024_report_full_extent", "WWF_PP_2024_report_full_extent.tif")

# Read in data
intact <- rast(intact_path)

# Crop + mask 
roi3_intact_crs <- project(roi3, crs(intact))
intact <- crop(intact, roi3_intact_crs)
intact <- mask(intact, roi3_intact_crs)

# Reclass and save to temp folder
intact_file <- file.path(tmp_dir, "intact_roi_rc.tif")

if (file.exists(intact_file)) {

  intact_rc <- rast(intact_file)
  
  } else {
    
    intact_rc <- ifel(
      intact == 0 | intact == 1,
      1,
      NA,
      filename = intact_file,
      overwrite = TRUE,
      datatype = "INT1U"   # 8-bit unsigned integer (small + fast)
    )
    
    }

# Convert to polygons
intact_smooth <- aggregate(intact_rc, fact = 2)
intact_poly <- as.polygons(intact_smooth, dissolve = TRUE, na.rm = TRUE)

# project vector 
intact_poly <- project(intact_poly, roi3)

# clean-up
rm(intact_path, intact, roi3_intact_crs, intact_rc, intact_smooth)
gc()
```

## 3.2 Native grasslands


```{r}
# Define path to the file(s)
native_path <- file.path(dat.dir, "3_grass_condition", "3.2_native_grass", "cfga_ngi_merged_reproj_algn_native.tif")

# Read in data
native <- rast(native_path)

# Crop 
roi3_native_crs <- project(roi3, crs(native))
native <- crop(native, roi3_native_crs)
native <- crop(native, roi3_native_crs)

# Convert to polygons
native_smooth <- aggregate(native, fact = 2)
native_poly <- as.polygons(native_smooth, dissolve = TRUE, na.rm = TRUE)

# Harmonize to Study Area CRS
native_poly <- project(native_poly, roi3)

# clean-up
rm(native_path, native, roi3_native_crs, native_smooth)
gc()
```

## 3.3 Connectivity

Current Density - Conductance?

```{r}
# Define path to the file(s)
conn_path <- file.path(dat.dir, "3_grass_condition", "3.3_connectivity", "Current_Density_1_Conductance.tif")

# Read in data
conn <- rast(conn_path)

# Harmonize to Study Area CRS
conn <- project(conn, roi3)

# Crop/mask before percentile calc
conn <- crop(conn, roi3, mask=T)

# Function to convert raster values to percentiles (0–1)
to_percentile <- function(r) {
  vals <- values(r, na.rm = TRUE)
  ecdf_fun <- ecdf(vals)
  
  r_pct <- r
  values(r_pct) <- ecdf_fun(values(r))
  
  return(r_pct)
}

# calc percentiles
conn_pct <- to_percentile(conn)

# threshold at 80th percentile
conn_top <- conn_pct >= 0.95

# convert to poly
conn_top_smooth <- aggregate(conn_top, fact = 2)
conn_poly <- as.polygons(conn_top_smooth, dissolve = TRUE, na.rm = TRUE)

# clean-up
rm(conn_path, conn, conn_pct, conn_top, conn_top_smooth)
gc()

```


# 4.0 Combine indicator data

```{r}
# special places
ind_sp <- c(kba, kba_draft, 
            priority, 
            comm_priority,
            rare1, rare2, rare3, rare4, rare5, rare6, rare7)

# species and habitats
ind_sh <- c(refugia_grassbird_poly, refugia_scrubbird_poly,
            upland,
            sar_top,
            farm_F_cap, farm_R_cap,
            farm_F_chng, farm_R_chng
            )

# grassland condition
ind_gc <- c(intact_poly,
            native_poly,
            conn_poly)


```


Crop to roi
```{r}
# function to crop svc
crop_svc_to_roi <- function(svc_obj, roi) {

  svc(
    lapply(as.list(svc_obj), function(x) {

      # Fix geometry if needed
      if (any(!is.valid(x), na.rm = TRUE)) {
        x <- makeValid(x)
      }

      # Use crop instead of intersect
      out <- crop(x, roi)

      return(out)
    })
  )
}

ind_sp_crop <- crop_svc_to_roi(ind_sp, roi3)
ind_sh_crop <- crop_svc_to_roi(ind_sh, roi3)
ind_gc_crop <- ind_gc # custom function fails on complex polys

```

# View the indicator shapefiles
WONKY LAYERS?
TBD

```{r}
plot_svc_layers <- function(svc_obj, roi = NULL, title_prefix = "") {

  layers <- as.list(svc_obj)

  for (i in seq_along(layers)) {

    x <- layers[[i]]

    plot(x,
         col = "lightgreen",
         border = "darkgreen",
         main = paste0(title_prefix, " layer ", i))

    if (!is.null(roi)) {
      plot(roi, add = TRUE, border = "red", lwd = 2)
    }
  }
}

# Run
plot_svc_layers(ind_sp_crop, roi3, "ind_sp_crop")
plot_svc_layers(ind_sh_crop, roi3, "ind_sh_crop")
plot_svc_layers(ind_gc_crop, roi3, "ind_gc_crop")


```



Save to s3
```{r}
# Special Places
numbers_sp <- c("1_1a", "1_1b", "1_2", "1_3",
                "1_4a", "1_4b", "1_4c", "1_4d", "1_4e", "1_4f", "1_4g")

stopifnot(length(numbers_sp) == length(ind_sp_crop))

for (i in seq_along(ind_sp_crop)) {

  fname <- paste0("ind_sp_", numbers_sp[i], ".gpkg")

  writeVector(
    ind_sp_crop[[i]],
    file.path(ind.dir, fname),
    overwrite = TRUE
  )
}

# Species + Habitats
numbers_sh <- c("2_1a", "2_1b",
                "2_2","2_3", 
                "2_4a", "2_4b", "2_5a", "2_5b")

stopifnot(length(numbers_sh) == length(ind_sh_crop))

for (i in seq_along(ind_sh_crop)) {

  fname <- paste0("ind_sh_", numbers_sh[i], ".gpkg")

  writeVector(
    ind_sh_crop[[i]],
    file.path(ind.dir, fname),
    overwrite = TRUE
  )
}

# Grassland Condition
numbers_gc <- c("3_1", "3_2", "3_3")

stopifnot(length(numbers_gc) == length(ind_gc_crop))

for (i in seq_along(ind_gc_crop)) {

  fname <- paste0("ind_gc_", numbers_gc[i], ".gpkg")

  writeVector(
    ind_gc_crop[[i]],
    file.path(ind.dir, fname),
    overwrite = TRUE
  )
}

```

