---
title: "Summarize Results"
output: # html_notebook
editor_options: 
  chunk_output_type: console
---

# R Set-up

```{r}
# Clear workspace
rm(list = ls())

# Libraries 
library(terra)
library(here)
library(purrr)
library(dplyr)
library(tigris)
library(sf)
library(ggplot2)
library(scales)

# Authenticate aws
source(here::here('aws_cred.R'))

# Base S3 location (GDAL-style path)
s3_base <- "/vsis3/sagemaker-gst-stage.sharing/ciara-hovis/TomoPrairie/bd_data"
s3_base_res <- "/vsis3/sagemaker-gst-stage.sharing/ciara-hovis/TomoPrairie"

dat.dir <- file.path(s3_base, "raw")
roi.dir <- file.path(s3_base, "studyarea")
dum.dir <- file.path(s3_base, "dummy")
ind.dir <- file.path(s3_base, "indicators")
results.dir <- file.path(s3_base_res, "results")


# Define directories (OLD)
# dir <- here()
# dat.dir <- file.path(dir, "data", "raw")
# roi.dir <- file.path(dir, "data", "studyarea")
# dum.dir <- file.path(dir, "data", "dummy")
```

We developed biodiversity indicators in three categories:  Special Places, Species & Habitats, and Grassland Condition. 


# 0.1 Study Area

Load in the study areas. Will align all data to this crs.

ROI 1 = TBD
ROI 2 = TBD
ROI 3 = Clip any data to this one

```{r}

roi1 <- vect(file.path(roi.dir, "class_study_area_aligned_cnvc_Dec2025.shp"))
roi2 <- vect(file.path(roi.dir, "modelling_extent_aligned_cnvc_Jan2026.shp"))
roi3 <- vect(file.path(roi.dir, "modelling_study_area_aligned_cnvc_Jan2026.shp"))

# plot(roi1)
# plot(roi2)
# plot(roi3, main="roi3")

```



## 0.2 Dummy Data

```{r}
dum_base <- rast(file.path(dum.dir, "eccc_reclass_reproj_algn_binary.tif"))
dum_base <- rast(file.path(dum.dir, "gpw_reproj_algn_natseminat_noshrub_masked.tif"))
dum_fut <- rast(file.path(dum.dir, "AAFC_LUTS_2020_5-classes.tif"))

# plot(dum_base)
# plot(dum_fut)

```

## 0.2 Eco-district Data
```{r}
# eco_dist$ECODISTRIC is ID column
eco_dist <-  vect(file.path(roi.dir, "national-ecological-framework-for-canada", "Ecodistricts",  "ecodistricts_clean_modelling-extent.shp"))

# plot(eco_dist)

eco_dist_lut <- s3read_using(
  read.csv,
  bucket = "sagemaker-gst-stage.sharing",
  object = "ciara-hovis/TomoPrairie/results/ecodistrict_lookup.csv"
)



```


## 0.3 Results Data

```{r}
ind_study_area <- s3read_using(
  read.csv,
  bucket = "sagemaker-gst-stage.sharing",
  object = "ciara-hovis/TomoPrairie/results/indicators_studyarea_pixel_counts.csv"
)

ind_ecodist <- s3read_using(
  read.csv,
  bucket = "sagemaker-gst-stage.sharing",
  object = "ciara-hovis/TomoPrairie/results/indicators_ecodistrict_pixel_counts.csv"
)

```

## 0.4 Province Data
```{r}
prov <-  vect(file.path(roi.dir, "can_provinces","lpr_000b21a_e.shp"))
prov <- project(prov, dum_base)

```



## 0.5 Indicator Data
```{r}
# Find matching GeoPackage files
objs <- get_bucket(
  bucket = "sagemaker-gst-stage.sharing",
  prefix = "ciara-hovis/TomoPrairie/bd_data/indicators/"
)

gpkg_files <- paste0(
  "/vsis3/sagemaker-gst-stage.sharing/",
  sapply(objs, `[[`, "Key")
)

gpkg_files <- gpkg_files[grepl("\\.gpkg$", gpkg_files)]

vectors <- lapply(gpkg_files, vect)

```

# 1.0 Summarize Tabular Results

```{r}
# Convert to area
ind_study_area$m2 <- ind_study_area$pixel_ones * 900
ind_study_area$ha <- ind_study_area$pixel_ones * 0.09

ind_ecodist$m2 <- ind_ecodist$pixel_ones * 900
ind_ecodist$ha <- ind_ecodist$pixel_ones * 0.09

```

Calc total grassland in study area
```{r}
# ---- Crop base raster ----
dum_base <- crop(dum_base, roi3, mask = TRUE)

# Count pixels equal to 1
n_cells_1 <- global(dum_base == 1, "sum", na.rm = TRUE)[1,1]

# Area in square meters
area_m2 <- n_cells_1 * 30 * 30

# Area in square meters
area_ha <- n_cells_1 * 0.09

area_m2
area_ha


```

Calc grass land by ecoregion
```{r}
# ---- Keep only grassland (value == 1) ----
grass <- dum_base == 1
eco_reg <-  vect(file.path(roi.dir, "national-ecological-framework-for-canada", "Ecoregions",  "ecoregions_clean_modelling-extent.shp"))


# Plot polygons colored by ECOREGION1
plot(
  eco_reg,
  "ECOREGION1",
  col = rainbow(length(unique(eco_reg$ECOREGION1))),
  border = "grey40",
  main = "Ecoregions"
)

plot(roi3, add=T, col="purple")
# ---- Extract grassland pixel counts by ecoregion ----
grass_by_eco <- extract(
  grass,
  eco_reg,
  fun = sum,
  na.rm = TRUE
)

# ---- Combine with ecoregion names ----
eco_grass_area <- data.frame(
  ECOREGION1   = eco_reg$ECOREGION1,
  EcoREgion_ID = eco_reg$ECOREGION_,
  grass_cells = grass_by_eco[, 2]
)

# ---- Convert pixel counts to area ----
# 30 m x 30 m = 900 m^2 = 0.09 ha
eco_grass_area$grass_m2 <- eco_grass_area$grass_cells * 900
eco_grass_area$grass_km2 <- (eco_grass_area$grass_cells * 900)/1000000
eco_grass_area$grass_ha <- eco_grass_area$grass_cells * 0.09

# ---- Final clean data frame ----
eco_grass_df <- as.data.frame(eco_grass_area)[, c("EcoREgion_ID", "ECOREGION1", "grass_m2", "grass_km2", "grass_ha")]

eco_grass_df

bucket_name <- "sagemaker-gst-stage.sharing"
eco_grass_object <- "ciara-hovis/TomoPrairie/results/eco_grass_area.csv"

tmp_file <- tempfile(fileext = ".csv")

write.csv(
  eco_grass_df,
  tmp_file,
  row.names = FALSE
)

put_object(
  file   = tmp_file,
  bucket = bucket_name,
  object = eco_grass_object
)

```


```{r}

ind_ecoreg <- ind_ecodist %>%
  left_join(
    eco_dist_lut,
    by = "ECODISTRIC"
  )

ind_ecoreg_sum <- ind_ecoreg %>%
  group_by(indicator_id, ECOREGION_) %>%
  summarise(
    ha = sum(ha, na.rm = TRUE),
    .groups = "drop"
  )

ind_ecozon_sum <- ind_ecoreg %>%
  group_by(indicator_id, ECOZONE_ID) %>%
  summarise(
    ha = sum(ha, na.rm = TRUE),
    .groups = "drop"
  )

bucket_name <- "sagemaker-gst-stage.sharing"
results_object <- "ciara-hovis/TomoPrairie/results/indicator_ecoregion_ha.csv"
results2_object <- "ciara-hovis/TomoPrairie/results/indicator_ecozone_ha.csv"

tmp_file <- tempfile(fileext = ".csv")

write.csv(
  ind_ecoreg_sum,
  tmp_file,
  row.names = FALSE
)

put_object(
  file   = tmp_file,
  bucket = bucket_name,
  object = results_object
)

tmp_file <- tempfile(fileext = ".csv")

write.csv(
  ind_ecozon_sum,
  tmp_file,
  row.names = FALSE
)

put_object(
  file   = tmp_file,
  bucket = bucket_name,
  object = results2_object
)

```



# 2.0 Make Maps

Baseline grassland
```{r}

library(terra)
library(sf)
library(ggplot2)
library(scales)
library(units)


# ---- Aggregate raster ----
dum_plot <- aggregate(dum_base, fact = 33, fun = mean, na.rm = TRUE)

grass_df <- as.data.frame(dum_plot, xy = TRUE, na.rm = TRUE)
names(grass_df)[3] <- "grass_prop"

# ---- Raster footprint (AFTER dum_plot exists) ----
raster_extent <- as.polygons(ext(dum_plot), crs = crs(dum_plot))
raster_extent_sf <- st_as_sf(raster_extent)

# ---- Saskatoon (projected correctly) ----
saskatoon <- vect(
  data.frame(
    name = "Saskatoon",
    lon = -106.6702,
    lat = 52.1579
  ),
  geom = c("lon", "lat"),
  crs = "EPSG:4326"
)

saskatoon <- project(saskatoon, crs(dum_plot))
saskatoon_df <- as.data.frame(saskatoon, geom = "xy")

# ---- Province borders (sf + intersection) ----
prov_sf <- st_as_sf(prov)

# Keep only provinces intersecting the grassland raster
prov_clip_sf <- prov_sf[
  st_intersects(prov_sf, raster_extent_sf, sparse = FALSE),
]

# ---- Study area ----
roi3_sf <- st_as_sf(roi3)

# ---- NeW: zoom extent based on study area + buffer ----
roi_bbox <- st_bbox(roi3_sf)

buffer_m <- 50000  # 50 km buffer (adjust if desired)

xlim <- c(roi_bbox["xmin"] - buffer_m,
          roi_bbox["xmax"] + buffer_m)

ylim <- c(roi_bbox["ymin"] - buffer_m,
          roi_bbox["ymax"] + buffer_m)

# ---- Plot ----
ggplot() +

  # Grassland raster
  geom_tile(
    data = grass_df,
    aes(x = x, y = y, fill = grass_prop)
  ) +

  # Intersecting provinces
  geom_sf(
    data = prov_clip_sf,
    fill = NA,
    color = "black",
    linewidth = 0.4
  ) +

  # Study area
  geom_sf(
    data = roi3_sf,
    fill = NA,
    color = "red",
    linewidth = 0.6
  ) +

  # Saskatoon point
  geom_point(
    data = saskatoon_df,
    aes(x = x, y = y),
    color = "red",
    size = 3
  ) +

  # Saskatoon label
  geom_text(
    data = saskatoon_df,
    aes(x = x, y = y, label = name),
    nudge_y = 40000,
    size = 4,
    fontface = "bold",
    color = "white"
  ) +

  scale_fill_viridis_c(
    option = "viridis",
    name = "Grassland cover",
    labels = percent_format(accuracy = 1),
    guide = guide_colorbar(
      barwidth = unit(16, "cm"),
      barheight = unit(0.6, "cm"),
      title.position = "top",
      label.position = "bottom"
    )
  ) +

  # Zoom to study area + buffer
  coord_sf(
    crs  = st_crs(roi3_sf),
    xlim = xlim,
    ylim = ylim,
    expand = FALSE
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  ) +
  labs(
    title = "Grassland Distribution",
    subtitle = "Proportion grassland per ~1 km² pixel"
  )

grassland_plot <- last_plot()

```


Indicators (study area)
```{r}
library(tidyterra)

# Convert indicator vectors to sf (REQUIRED)
vectors_sf <- lapply(vectors, st_as_sf)

# Make sure ROI and provinces are sf (safe even if already sf)
roi3_sf <- st_as_sf(roi3)
prov_clip_sf <- st_as_sf(prov_clip_sf)

ggplot() +

  # Grassland raster (FAST, native)
  geom_spatraster(
    data = dum_base
    ) +
  
  scale_fill_gradient(
  low = "white",
  high = "gray26",
  na.value = "transparent",
  guide = "none"
  ) +

  # Province borders
  geom_sf(
    data = prov_clip_sf,
    fill = NA,
    color = "black",
    linewidth = 0.4
    ) +

  # Indicator vectors
  geom_sf(
    data = vectors_sf[[1]],
    aes(color = "KBAs"),
    fill = alpha("#4477AA", 0.3),
    linewidth = 0.6
    ) +
  
  geom_sf(
    data = vectors_sf[[2]],
    aes(color = "KBAs (draft)"),
    fill = alpha("#228833", 0.3),
    linewidth = 0.6
    ) +
  
  geom_sf(
    data = vectors_sf[[3]],
    aes(color = "ECCC for SAR"),
    fill = alpha("#CCBB44", 0.3),
    linewidth = 0.6
  ) +
  geom_sf(
    data = vectors_sf[[4]],
    aes(color = "ECCC Comm Nom"),
    fill = alpha("#AA3377", 0.3),
    linewidth = 0.6
  ) +

  # Study area
  geom_sf(
    data = roi3_sf,
    fill = NA,
    color = "red",
    linewidth = 0.8
  ) +

 scale_color_manual(
  breaks = c(
    "KBAs",
    "KBAs (draft)",
    "ECCC for SAR",
    "ECCC Comm Nom"
  ),
  values = c(
    "KBAs" = "#4477AA",
    "KBAs (draft)" = "#228833",
    "ECCC for SAR" = "#CCBB44",
    "ECCC Comm Nom" = "#AA3377"
  ),
  name = "Indicators"
) +

  coord_sf(
    crs = st_crs(roi3_sf),
    xlim = c(st_bbox(roi3_sf)["xmin"] - 50000,
             st_bbox(roi3_sf)["xmax"] + 50000),
    ylim = c(st_bbox(roi3_sf)["ymin"] - 50000,
             st_bbox(roi3_sf)["ymax"] + 50000),
    expand = FALSE
  ) +

  theme_minimal(base_size = 14) +
  
  theme(
    panel.grid = element_blank(),
    axis.text  = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  ) +
  
  labs(
    title = "Biodiversity Indicators",
    subtitle = "Grassland Area (Grey)"
  )

indicators_plot <- last_plot()

```


Eco-district plots
```{r}
# Make chlorpleth maps of ecodistricts

# Convert ecodistrict polygons to sf
eco_sf <- st_as_sf(eco_dist)

# Join indicator summary to polygons
eco_ind_sf <- eco_sf %>%
  left_join(
    ind_ecodist,
    by = c("ECODISTRIC" = "ECODISTRIC")
  )

indicator_labs <- c(
  "1" = "KBAs",
  "2" = "KBAs (draft)",
  "3" = "ECCC for SAR",
  "4" = "ECCC Comm Nom"
)

indicator_cols <- c(
  "1" = "#4477AA",  # blue
  "2" = "#228833",  # green
  "3" = "#CCBB44",  # yellow
  "4" = "#AA3377"   # pink
)

eco_ind_sf$indicator_f <- factor(
  eco_ind_sf$indicator_id,
  levels = names(indicator_labs),
  labels = indicator_labs
)


eco_sf <- st_as_sf(eco_dist)

eco_ind_complete <- eco_sf %>%
  left_join(
    ind_ecodist,
    by = "ECODISTRIC"
  ) %>%
  mutate(
    ha = ifelse(is.na(ha), 0, ha)
  )

```

```{r}

make_indicator_map <- function(ind_id, base_col, title_txt) {

  ind_dat <- eco_ind_complete[
    eco_ind_complete$indicator_id == ind_id,
  ]

  ggplot() +

    geom_sf(
      data = ind_dat,
      aes(fill = ha),
      color = NA
    ) +

    geom_sf(
      data = eco_sf,
      fill = NA,
      color = "grey70",
      linewidth = 0.2
    ) +
    
    geom_sf(
      data = roi3_sf,
      fill = NA,
      color = "red",
      linewidth = 0.2
    ) +

    
    scale_fill_gradientn(
      colours = c(
        "#F2F2F2",        # light neutral grey (zero)
        base_col
      ),
      values = c(0, 1),
      labels = scales::comma,
      name = paste0(title_txt, " area (ha)"),
      guide = guide_colorbar(
        barwidth  = unit(10, "cm"),
        barheight = unit(0.6, "cm"),
        title.position = "top",
        label.position = "bottom"
      )
    ) +

    coord_sf(expand = FALSE) +

    theme_minimal(base_size = 12) +
    theme(
      panel.grid   = element_blank(),
      axis.text    = element_blank(),
      axis.title   = element_blank(),
      axis.ticks   = element_blank(),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )
}




map_kba <- make_indicator_map(
  ind_id   = 1,
  base_col = "#4477AA",
  title_txt = "KBAs"
)

map_kba_draft <- make_indicator_map(
  ind_id   = 2,
  base_col = "#228833",
  title_txt = "KBAs (draft)"
)

map_sar <- make_indicator_map(
  ind_id   = 3,
  base_col = "#CCBB44",
  title_txt = "ECCC for SAR"
)

map_comm <- make_indicator_map(
  ind_id   = 4,
  base_col = "#AA3377",
  title_txt = "ECCC Comm Nom"
)

library(patchwork)

combined_ecodist_plot <-
  (map_kba | map_kba_draft) /
  (map_sar | map_comm) +
  plot_annotation(
    title = "Indicator Coverage by Eco‑District",
    subtitle = "Total indicator area (hectares)",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16)
    )
  )

combined_ecodist_plot
```

Save plots
```{r}

fig_width  <- 13.33  # inches (16:9 slide)
fig_height <- 7.5
fig_dpi    <- 300

bucket_name <- "sagemaker-gst-stage.sharing"

grassland_obj  <- "ciara-hovis/TomoPrairie/figures/Baseline_grassland.png"
indicators_obj <- "ciara-hovis/TomoPrairie/figures/Indicators_study_area.png"
ecodist_ind_obj <- "ciara-hovis/TomoPrairie/figures/Indicators_ecodist.png"

grassland_file <- tempfile(fileext = ".png")
indicators_file <- tempfile(fileext = ".png")
ecodist_ind_file <- tempfile(fileext = ".png")

ggsave(
  filename = grassland_file,
  plot     = grassland_plot,
  width    = fig_width,
  height   = fig_height,
  units    = "in",
  dpi      = fig_dpi,
  bg       = "white"
)
ggsave(
  filename = indicators_file,
  plot     = indicators_plot,
  width    = fig_width,
  height   = fig_height,
  units    = "in",
  dpi      = fig_dpi,
  bg       = "white"
)
ggsave(
  filename = ecodist_ind_file,
  plot     = combined_ecodist_plot,
  width    = fig_width,
  height   = fig_height,
  units    = "in",
  dpi      = fig_dpi,
  bg       = "white"
)

put_object(
  file   = grassland_file,
  bucket = bucket_name,
  object = grassland_obj
)
put_object(
  file   = indicators_file,
  bucket = bucket_name,
  object = indicators_obj
)
put_object(
  file   = ecodist_ind_file,
  bucket = bucket_name,
  object = ecodist_ind_obj
)

```



